Reverse eigenvalue problem

### 2024-03-08 Docs/NhanesApplExample.Rmd
weekly update

---
title: "Problem of Step 4 (debias) of fGFPCA on NHANES data"
author: "Ying Jin"
output: 
  html_document:
    self_contained: yes
    number_sections: true
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: true
    font: 12pt
---

@@

set.seed(227)

@@

```{r}
# load data
df <- read_rds(here("Data/nhanes_bi.rds"))

# code
source(here("Code/GLMM-FPCA.R")) 
# use pred_latent function to estimate latent function 
K <- 4
source(here("Code/OutsampBayes.R"))
```

```{r bin_data}
# head(df)
df <- df %>% rename(id=SEQN, Y=Z)
N <- length(unique(df$id))
Ntr <- 500
J <- length(unique(df$sind))

load(here("Data/Appl_debias_model.RData"))

# parameters used in binning
bin_w <- 10 # bin width
n_bin <- J/bin_w # number of bins
brks <- seq(0, J, by = bin_w) # cutoff points
mid <- (brks+bin_w/2)[1:n_bin] # mid points
```

# Estimates

- Eigenvalues

```{r evals}
evals <- 1/debias_glmm$sp[2:5]/n_bin
```

- PC functions

```{r}
df_phi <- debias_glmm$model %>% select(sind, starts_with("phi")) %>% distinct(.) %>%
  mutate_at(vars(starts_with("phi")), function(x){x*sqrt(n_bin)})

df_phi %>% pivot_longer(2:5) %>%
  ggplot(aes(x=sind, y=value))+
  geom_line()+
  facet_wrap(~name)
```

- Mean function

```{r}
mu <- predict(debias_glmm, type = "terms")[1:J, 1]
# plot(1:J, mu, type = "l", main = "Population mean")
plot(1:J, plogis(mu), type = "l", main = "Population mean on the probablity scale")
```

- Deviation from each PC from the population mean

Below are figures for $\hat{f}_0(t)+2\sqrt{\hat{\lambda}}_k\phi_l(k)$, but on the probablity scale

```{r}
df_pc_dev <- df_phi %>%
  mutate(mu = mu) %>%
  mutate(phi1_dev = mu + 2*sqrt(evals[1])*phi1,
         phi2_dev = mu + 2*sqrt(evals[2])*phi2,
         phi3_dev = mu + 2*sqrt(evals[3])*phi3,
         phi4_dev = mu + 2*sqrt(evals[4])*phi4)

df_pc_dev %>%
  pivot_longer(7:10) %>%
  mutate(value = plogis(value), mu = plogis(mu)) %>%
  ggplot()+
  geom_line(aes(x=sind, y=value))+
  geom_line(aes(x=sind, y=mu), col = "red")+
  facet_wrap(~name) 
```

It looks like subjects with high loadings on PC2 would be more active than population average during afternoon/night, and less active in the morning. PC3, on the other hand, seems to indicate more active at night and early morning and less active in midday. They both reflect more active at night time.

- Individual score

```{r}
data.frame(matrix(debias_glmm$coefficients[-c(1:9)], ncol = 4)) %>%
  pivot_longer(1:4) %>%
  ggplot(aes(x=name, y=value))+
  geom_boxplot()+
  geom_jitter(size = 0.2, alpha = 0.2)
```

Compared to the other three PCs, the distribution of $\hat{\xi}_{i3}$ seems to be shifted lower (positively skewed?).

# Simulation using re-evaluated estimates

```{r gen_data}
# generate score
xi <- rmvnorm(Ntr, mean = rep(0, K), sigma = diag(evals))
# xi %>% data.frame() %>% pivot_longer(1:4, names_to = "PC", values_to = "score") %>%
#   ggplot()+
#   geom_boxplot(aes(x=PC, y=score))+
#   geom_jitter(aes(x=PC, y=score), size = 0.5)

@@

# add in mean 
eta <- apply(bi, 2, function(x){x+mu})
gen_df <- data.frame(id = factor(rep(1:Ntr, each=J)),
                     sind = rep(1:J, Ntr), 
                     eta_i = as.vector(eta))

# binary outcome
gen_df$Y <- rbinom(Ntr*J, size=1, prob=plogis(gen_df$eta_i))

gen_df %>% filter(id %in% 1:4) %>%
  ggplot()+
  geom_point(aes(x=sind, y=Y), size = 0.5)+
  geom_line(aes(x=sind, y=plogis(eta_i)), col = "red")+
  facet_wrap(~id)+
  labs(title = "Generated data")
```

# fGFPCA on generated data

```{r fGFPCA1}
# bin 
gen_df$bin <- cut(gen_df$sind, breaks = brks, include.lowest = T, labels = mid)
gen_df$bin <- as.numeric(as.character(gen_df$bin))

# local GLMM
train_bin_lst <- split(gen_df, f = gen_df$bin)
df_est_latent <- lapply(train_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
df_est_latent <- bind_rows(df_est_latent)

df_est_latent %>% filter(id %in% 1:4) %>%
  ggplot()+
  geom_point(aes(x=sind, y=Y), size = 0.5)+
  geom_line(aes(x=sind, y=plogis(eta_i), col = "True"))+
  geom_line(aes(x=sind, y=plogis(eta_hat), col = "Estimate"))+
  facet_wrap(~id)+
  labs(title = "Generated data")
```

```{r fGFPCA2}
# FPCA
uni_eta_hat <- df_est_latent %>% filter(bin==sind)
mat_est_unique <- matrix(uni_eta_hat$eta_hat, nrow=Ntr, ncol=n_bin, byrow = F) 
fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T, npc=K)
```

```{r re_eval_PC}
# interpolation
p <- 3 # order of b splines 
knots <- 35 # number of knots (same from FPCA model)
knots_values <- seq(-p, knots + p, length = knots + 1 + 2 *p)/knots
knots_values <- knots_values * (max(mid) - min(mid)) + min(mid)

est_phi <- matrix(NA, J, K) 
for(k in 1:K){
  lm_mod <- lm(fpca_mod$efunctions[,k] ~ B-1)
  est_phi[,k] <- Bnew %*% coef(lm_mod)
} 
colnames(est_phi) <- paste0("phi", 1:K)
```

I would like to look at estimated PC functions:

```{r}
theme_set(theme_minimal())
cols <- c(brewer.pal(4, "Set2"), "#000000") # define a color palette 
names(cols) <- c("0.2", "0.4", "0.6", "0.8", "True")

# PC functions after re-evaluation
bind_rows(
  data.frame(sind=1:J, est_phi*sqrt(n_bin), type = "Simulation"),
  data.frame(df_phi, type = "Generation")
) %>%
  rename("PC1" = phi1, "PC2" = phi2, "PC3" = phi3, "PC4" = phi4) %>%
  pivot_longer(2:5) %>%
  ggplot()+
  geom_line(aes(x=sind, y=value, col=type), linewidth=1)+
  facet_wrap(~name)+
  scale_color_brewer(palette = "Set2")+
   scale_x_continuous(breaks = seq(0, 1440, by = 360),
                     labels = c("0am", "6am", "12pm", "6pm", "12am"))+
  labs(x = "Time", y="", color = "")
```

It looks like something went wrong in the estimation of PC3. The estimated line is either identical or off by a sign compared to the true PC functions. That said, PC2 seems a bit off too.

What about the estimated PC functions from FPCA without re-evaluation?

```{r}
fpca_phi <- fpca_mod$efunctions*sqrt(n_bin)
colnames(fpca_phi) <- paste0("phi", 1:K)
bind_rows(
  data.frame(df_phi, type = "True"),
  data.frame(sind = mid, fpca_phi, type = "FPCA")
) %>%
  pivot_longer(2:5) %>%
  ggplot()+
  geom_point(aes(x=sind, y=value, col=type),size = 0.5)+
  facet_wrap(~name)
```

Next I'd like to re-evaluate the eigenvalues.

PS from the mgcv [user manual](https://cran.r-project.org/web/packages/mgcv/mgcv.pdf): *Note that discrete=TRUEmay result in re-ordering of variables in tensor product smooths for improved efficiency, and sp must be
supplied in re-ordered order*. Does this also apply to random effects? Does it have anything to do with the reversed eigenvalue functions? (I don't believe this is the case though, based on the experiments. Basciall I fit the model on a subset of 100 subjects setting discrete = F, and the smoothing parameters still flipped).

```{r}
# debias model
gen_df <- gen_df %>% 
  left_join(df_phi, by = "sind")
gen_df$id <- as.factor(gen_df$id)
debias_glmm2 <- bam(Y ~ s(sind, bs="cr")+
                   s(id, by=phi1, bs="re")+
                   s(id, by=phi2, bs="re")+
                   s(id, by=phi3, bs="re")+
                   s(id, by=phi4, bs="re"),
                 family = binomial,
                 data=gen_df,
                 method = "fREML",
                 discrete = TRUE)
```

- Eigenvalues before and after re-evaluation

```{r, class.source='fold-show'}
# fpca
fpca_mod$evalues/n_bin

# re-evaluated
1/debias_glmm2$sp[2:5]
```

It looks like the flip happened again.

```{r, eval=FALSE}
new_df <- gen_df %>% filter(id == 1)
est_phi <- predict(debias_glmm2, newdata = new_df, type = "terms")[, 2:5] %>% 
  data.frame()
est_phi %>%
  mutate(sind=1:J) %>%
  pivot_longer(1:4) %>%
  ggplot()+
  geom_line(aes(x=sind, y= value))+
  facet_wrap(~name)
```

```{r eval=FALSE}
coef1 <- coef(debias_glmm2)[c("s(id):phi1.1", "s(id):phi2.1", "s(id):phi3.1", "s(id):phi4.1")]
coef1 <- matrix(coef1, nrow = J, ncol = 4, byrow = T)
(as.matrix(df_phi[, 2:5]) * coef1) %>% data.frame() %>%
  mutate(sind=1:J) %>%
  pivot_longer(1:4) %>%
  ggplot()+
  geom_line(aes(x=sind, y= value))+
  facet_wrap(~name)

@@

At this point, I think we could say the reversed eigenvalue problem is an artifact of the dataset it self. It looks like FPCA captures the shape of eigenvalues very well, but for some reason, mgcv::bam thinks it is necessary to penalize the second eigenfunctions more than the third function. Why is that? Could it be that PC2 is more wiggly?

If we compare the eigenvalues used for data generation to the ones estimated by bam, we can see bam actually does a good job. But the order of eigenfunctions flipped, and that is because FPCA re-ordered the eigenfunctions based on eigenvalues.

```{r, class.source='fold-show'}
# generation
evals
# estimation
1/debias_glmm2$sp[2:5]
```

>>>>

---
title: "Reverse eigenvalue problem"
author: "Ying Jin"
output: 
  html_document:
    self_contained: yes
    number_sections: true
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: true
    font: 12pt
---

@@

set.seed(1114)

@@

```{r load_data}
# load data
df <- read_rds(here("Data/nhanes_bi.rds"))
df <- df %>% rename(id=SEQN, Y=Z)

N <- length(unique(df$id))
J <- length(unique(df$sind))
K <- 4 # number of eigenfunctions to use
```

```{r load_code}
source(here("Code/GLMM-FPCA.R")) 
source(here("Code/OutsampBayes.R"))
```

It seems that the reversed eigenvalue problem has to do with 1) the choice of spline basis and 2) sample size.

# Spline basis

I'll take a subset of 500 subject and try different spline basis.

```{r data_split}
# or for sanity checks, use a subset of 1000 subjects
Ntry <- 500
train_id <- sample(unique(df$id), size = Ntry)
train_df <- df %>% filter(id %in% train_id)
```

```{r set_up}
# basis to try
bs_vec <- c("cr", "cs", "tp", "bs", "ps")
evals <- matrix(NA, length(bs_vec), K)
```

```{r fGFPCA}
# Step 1: Bin every 10 observations
bin_w <- 10 # bin width
n_bin <- J/bin_w # number of bins
brks <- seq(0, J, by = bin_w) # cutoff points
mid <- (brks+bin_w/2)[1:n_bin] # mid points

train_df$bin <- cut(train_df$sind, breaks = brks, include.lowest = T, labels = mid)
train_df$bin <- as.numeric(as.character(train_df$bin))

# Step 2:  Local GLMM
train_bin_lst <- split(train_df, f = train_df$bin)
df_est_latent <- lapply(train_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
df_est_latent <- bind_rows(df_est_latent)

# FPCA
uni_eta_hat <- df_est_latent %>% filter(bin==sind)
mat_est_unique <- matrix(uni_eta_hat$eta_hat,
                         nrow=length(train_id), 
                         ncol=n_bin, byrow = F) 
fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T, npc=K) # keep 4 PCs

# Re-evaluation
## grid extension
p <- 3 # order of b splines 
knots <- 35 # number of knots (same from FPCA model)
knots_values <- seq(-p, knots + p, length = knots + 1 + 2 *p)/knots
knots_values <- knots_values * (max(mid) - min(mid)) + min(mid)

df_phi <- matrix(NA, J, K) 
for(k in 1:K){
  lm_mod <- lm(fpca_mod$efunctions[,k] ~ B-1)
  df_phi[,k] <- Bnew %*% coef(lm_mod)
}# project binned eigenfunctions onto the original grid

## debias
## let's try scale the eigenfunctions before re-evaluation
df_phi <- data.frame(sind = 1:J, df_phi)
colnames(df_phi) <- c("sind", paste0("phi", 1:K))
# df_phi <- df_phi %>% mutate(
#   phi1 = sqrt(n_bin)*phi1,
#   phi2 = sqrt(n_bin)*phi2,
#   phi3 = sqrt(n_bin)*phi3,
#   phi4 = sqrt(n_bin)*phi4
# )
train_df <- train_df %>% 
  left_join(df_phi, by = "sind")
train_df$id <- as.factor(train_df$id)

```{r try_splines, results='hide'}
pb <- txtProgressBar(min=0, max=length(bs_vec), style = 3)

t1 <- Sys.time()
for(i in seq_along(bs_vec)){

this_glmm <- bam(Y ~ s(sind, bs = bs_vec[i])+
                     s(id, by=phi1, bs="re")+
                     s(id, by=phi2, bs="re")+
                     s(id, by=phi3, bs="re")+
                     s(id, by=phi4, bs="re"),
                   family = binomial,
                   data=train_df,
                   method = "fREML",
                   discrete = TRUE)
evals[i, ] <- 1/this_glmm$sp[2:5]/n_bin
setTxtProgressBar(pb, i)

}

close(pb)
t2 <- Sys.time()
# t2-t1 # about 5 minutes
```

```{r}
data.frame(bs_vec, evals)
```

It seems the flip happened for both cubic spline basis sets, but not the other ones. Also, when sample size gets larger, the flip happens almost certainly for all spline basis.

When I increase the number of PC from 4 to 5, the second and third eigenvalues still flipped almost all the time.

```{r, eval=FALSE}
bind_rows(
data.frame(sind=1:J, mu = predict(this_glmm, type="terms")[1:J, 1]) %>% mutate(type = "BAM"),
data.frame(sind=mid, mu = fpca_mod$mu) %>% mutate(type = "FPCA")
) %>%
  ggplot()+
  geom_line(aes(x=sind, y=mu, col=type))+
  labs(title="Population mean estimate")

# Simulation using estimates from NHANES data

## Estimates use for simualtion

For now I will use estimates from the model fit on the whole NHANES dataset, even though the spline basis is not proper (bs = "cc").

```{r load_est}
load(here("Data/Appl_debias_model.RData"))

# eigenvalues
true_evals <- 1/debias_glmm$sp[2:5]/n_bin

# mean 
true_mu <- predict(debias_glmm, type = "terms")[, 1]
true_mu <- matrix(true_mu, nrow = J)[, 1]
# plot(1:J, true_mu)

# eigenfunctions
df_phi <- debias_glmm$model %>% select(sind, starts_with("phi")) %>% distinct(.) %>% 
  mutate_at(vars(starts_with("phi")), function(x){x*sqrt(n_bin)})
```

Below are some visualizations of these estimates

```{r plot_pc_func}
df_phi %>%
  pivot_longer(starts_with("phi")) %>%
  ggplot()+
  geom_line(aes(x=sind, y=value))+
  facet_wrap(~name)+
  labs(title = "Eigenfunctions")
```

```{r plot_pc_dev_from_mean}
df_pc_dev <- bind_rows(
  data.frame(sind = 1:J, mu=true_mu, pc = "PC1",
             high = true_mu + 2*sqrt(true_evals[1])*df_phi$phi1,
             low = true_mu - 2*sqrt(true_evals[1])*df_phi$phi1),
  data.frame(sind = 1:J, mu=true_mu, pc = "PC2",
             high = true_mu + 2*sqrt(true_evals[2])*df_phi$phi2,
             low = true_mu - 2*sqrt(true_evals[2])*df_phi$phi2),
  data.frame(sind = 1:J, mu=true_mu, pc = "PC3",
             high = true_mu + 2*sqrt(true_evals[3])*df_phi$phi3,
             low = true_mu - 2*sqrt(true_evals[3])*df_phi$phi3),
  data.frame(sind = 1:J, mu=true_mu, pc = "PC4",
             high = true_mu + 2*sqrt(true_evals[4])*df_phi$phi4,
             low = true_mu - 2*sqrt(true_evals[4])*df_phi$phi4)
)

# plot
df_pc_dev %>%
  ggplot()+
  geom_ribbon(aes(x=sind, y=mu, ymin=low, ymax=high, alpha = 0.2))+
  geom_line(aes(x=sind, y=mu))+
  facet_wrap(~pc)+
  labs(title = "Deviation of each PC from population mean")
```

```{r plot_score}
data.frame(matrix(debias_glmm$coefficients[-c(1:9)], ncol = 4)) %>%
  pivot_longer(1:4) %>%
  ggplot(aes(x=name, y=value))+
  geom_boxplot()+
  geom_jitter(size = 0.2, alpha = 0.2)+
  stat_summary(fun=mean, geom="point", color="red")+
  labs(title = "Individual scores")
```

## Data generation

```{r gen_data_unbound}
# interpolate FPCA mean
# interp_mu_fpca <- approx(mid, fpca_mod$mu, xout=1:J, rule=2)
# true_mu <- interp_mu_fpca$y

# increase sample size?
Ntry <- 500
# generate score
xi <- rmvnorm(Ntry, mean = rep(0, K), sigma = diag(true_evals))

@@

# add in unbounded mean
eta <- apply(bi, 2, function(x){x+true_mu})
gen_df <- data.frame(id = factor(rep(1:Ntry, each=J)),
                     sind = rep(1:J, Ntry),
                     eta_i = as.vector(eta))

# binary outcome
gen_df$Y <- rbinom(Ntry*J, size=1, prob=plogis(gen_df$eta_i))
```

```{r plot_data}
# comparing generated data to the actual data
bind_rows(
    gen_df %>% 
      group_by(sind) %>% 
      summarize(active = mean(Y)) %>%
      mutate(type="Simulation"),

    df %>% 
      group_by(sind) %>% 
      summarize(active = mean(Y)) %>%
      mutate(type="NHANES")
) %>%
  ggplot()+
      geom_line(aes(x=sind, y=active, col = type))+
      labs(title="Proportion of active subjects by minute")

@@

This is the first discrepancy I have found - It looks like the simulated data is more active and more variable than the real data. Doesn't that mean the algorithm overestimates the population activity and variability? Since the mean estimates will surely change in Step 4 (debias), would the overestimation be caused by that?

## fGFPCA

```{r exp_gen_data}
# bin data
# using the same bin length 
gen_df$bin <- cut(gen_df$sind, breaks = brks, include.lowest = T, labels = mid)
gen_df$bin <- as.numeric(as.character(gen_df$bin))

# Step 2:  Local GLMM
gen_bin_lst <- split(gen_df, f = gen_df$bin)
df_est_latent <- lapply(gen_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
df_est_latent <- bind_rows(df_est_latent)
## Figure
# df_est_latent %>% filter(id %in% 1:4) %>%
#   ggplot()+
#   geom_point(aes(x=sind, y=Y), size = 0.5)+
#   geom_line(aes(x=sind, y=plogis(eta_i), col = "true"))+
#   geom_line(aes(x=sind, y=plogis(eta_hat), col = "estimate"))+
#   facet_wrap(~id)

# FPCA
uni_eta_hat <- df_est_latent %>% filter(bin==sind)
mat_est_unique <- matrix(uni_eta_hat$eta_hat,
                        nrow=Ntry, 
                        ncol=n_bin, byrow = F) 
# dim(mat_est_unique)
fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T, npc=K) # keep 4 PCs
```

```{r}
# Re-evaluation
est_phi <- matrix(NA, J, K) 
p <- 3 # order of b splines 
knots <- 35 # number of knots (same from FPCA model)
knots_values <- seq(-p, knots + p, length = knots + 1 + 2 *p)/knots
knots_values <- knots_values * (max(mid) - min(mid)) + min(mid)

B <- spline.des(knots = knots_values, x = mid, ord = p + 1,
                outer.ok = TRUE)$design  # evaluate B-splines on binned grid
Bnew <- spline.des(knots = knots_values, x = 1:J, ord = p + 1,
                   outer.ok = TRUE)$design
for(k in 1:K){
  lm_mod <- lm(fpca_mod$efunctions[,k] ~ B-1)
  est_phi[,k] <- Bnew %*% coef(lm_mod)
}# project binned eigenfunctions onto the original grid

## debias
est_phi <- data.frame(sind = 1:J, est_phi)
colnames(est_phi) <- c("sind", paste0("phi", 1:K))
gen_df <- gen_df %>% left_join(est_phi, by = "sind")
gen_df$id <- as.factor(gen_df$id)
### model fit with B-splines basis
t1 <- Sys.time()
this_glmm <- bam(Y ~ s(sind, bs = "cr")+
                   s(id, by=phi1, bs="re")+
                   s(id, by=phi2, bs="re")+
                   s(id, by=phi3, bs="re")+
                   s(id, by=phi4, bs="re"),
                 family = binomial,
                 data=gen_df,
                 method = "fREML",
                 discrete = TRUE)
t2 <- Sys.time()
```

Below I compared the estimated eigenvalues from simulated data to the true eigenvalues. As we see, the reverse did not happen on the simulated data.

```{r new_evals, class.source='fold-show'}
# estimated eigenvalues
1/this_glmm$sp[2:5]/n_bin

# true eigenvalues
true_evals
```
Below is the comparision between true and esimtated eigenfunctions.

```{r new_PC}
bind_rows(df_phi %>% mutate(type = "True"),
          est_phi %>% mutate(type = "Estimate",
                             phi1 = phi1*sqrt(n_bin),
                             phi2 = phi2*sqrt(n_bin),
                             phi3 = phi3*sqrt(n_bin),
                             phi4 = phi4*sqrt(n_bin))) %>%
  pivot_longer(2:5) %>%
  ggplot()+
  geom_line(aes(x=sind, y=value, col = type))+
  facet_wrap(~name)
```

And below is the comparision between true mean, mean estimated from FPCA, and re-evaluated mean from the BAM model in Step 4. The re-evaluation seem to drive the mean function closer to the truth. But in that case, why would generated data much more active compared to the NHANES data?

```{r compare_mean}
bind_rows(
  data.frame(sind=1:J, mu=true_mu) %>% mutate(type = "True"),
  data.frame(sind=1:J, mu=predict(this_glmm, type = "terms")[1:J, 1]) %>% mutate(type = "BAM"),
  data.frame(sind=mid, mu=fpca_mod$mu) %>% mutate(type="FPCA")
) %>%
  ggplot()+
      geom_line(aes(x=sind, y=mu, col = type))+
      labs(title="Mean estimates")

# Simulation with the FPCA mean

I would like to experiments if I could simulated data whose activity level and variation is closer to NHANES, using the interpolated FPCA mean and un-evaluated eigenvalues.

```{r load_fpca_mu}
# load FPCA model from NHANES
load(here("Data/Appl_fpca_model.RData"))

# compare means on NHANES data from FPCA and debias model
bind_rows(
  data.frame(sind=1:J, mu = true_mu, type = "BAM"),
  data.frame(sind=mid, mu = fpca_mod$mu, type = "FPCA")
) %>% 
  ggplot()+
  geom_line(aes(x=sind, y=mu, col = type))+
  labs(title = "Mean estimates on the NHANES dataset")

## Data generation

```{r gen_data_fpca_mu}
# interpolate FPCA mean
interp_mu_fpca <- approx(mid, fpca_mod$mu, xout=1:J, rule=2)
true_mu <- interp_mu_fpca$y

# FPCA eigenvalues
true_evals <- fpca_mod$evalues/n_bin

# maybe increase sample size?
# Ntry <- 1000

# generate score
xi <- rmvnorm(Ntry, mean = rep(0, K), sigma = diag(true_evals))

# add in un-evaluated mean
eta <- apply(bi, 2, function(x){x+true_mu})
gen_df <- data.frame(id = factor(rep(1:Ntry, each=J)),
                     sind = rep(1:J, Ntry),
                     eta_i = as.vector(eta))

# binary outcome
gen_df$Y <- rbinom(Ntry*J, size=1, prob=plogis(gen_df$eta_i))
```

```{r plot_data_fpca_mu}
# comparing generated data to the actual data
bind_rows(
    gen_df %>% 
      group_by(sind) %>% 
      summarize(active = mean(Y)) %>%
      mutate(type="Simulation"),

    df %>% 
      group_by(sind) %>% 
      summarize(active = mean(Y)) %>%
      mutate(type="NHANES")
) %>%
  ggplot()+
      geom_line(aes(x=sind, y=active, col = type))+
      labs(title="Proportion of active subjects by minute")

Looks like by using the FPCA mean, I was able to lower the activity level of generated data down to the level of NHANES data.

## fGFPCA

```{r fgfpca_fpca_mean}
# bin data
# using the same bin length 
gen_df$bin <- cut(gen_df$sind, breaks = brks, include.lowest = T, labels = mid)
gen_df$bin <- as.numeric(as.character(gen_df$bin))

# Step 2:  Local GLMM
gen_bin_lst <- split(gen_df, f = gen_df$bin)
df_est_latent <- lapply(gen_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
df_est_latent <- bind_rows(df_est_latent)
## Figure
# df_est_latent %>% filter(id %in% 1:4) %>%
#   ggplot()+
#   geom_point(aes(x=sind, y=Y), size = 0.5)+
#   geom_line(aes(x=sind, y=plogis(eta_i), col = "true"))+
#   geom_line(aes(x=sind, y=plogis(eta_hat), col = "estimate"))+
#   facet_wrap(~id)

# FPCA
uni_eta_hat <- df_est_latent %>% filter(bin==sind)
mat_est_unique <- matrix(uni_eta_hat$eta_hat,
                        nrow=Ntry, 
                        ncol=n_bin, byrow = F) 
# dim(mat_est_unique)
fpca_mod_sim <- fpca.face(mat_est_unique, argvals = mid, var=T, npc=K) # keep 4 PCs
```

```{r}
# Re-evaluation
est_phi <- matrix(NA, J, K) 
p <- 3 # order of b splines 
knots <- 35 # number of knots (same from FPCA model)
knots_values <- seq(-p, knots + p, length = knots + 1 + 2 *p)/knots
knots_values <- knots_values * (max(mid) - min(mid)) + min(mid)

B <- spline.des(knots = knots_values, x = mid, ord = p + 1,
                outer.ok = TRUE)$design  # evaluate B-splines on binned grid

Bnew <- spline.des(knots = knots_values, x = 1:J, ord = p + 1,
                   outer.ok = TRUE)$design
for(k in 1:K){
  lm_mod <- lm(fpca_mod_sim$efunctions[,k] ~ B-1)
  est_phi[,k] <- Bnew %*% coef(lm_mod)
}# project binned eigenfunctions onto the original grid

## debias
est_phi <- data.frame(sind = 1:J, est_phi)
colnames(est_phi) <- c("sind", paste0("phi", 1:K))
gen_df <- gen_df %>% left_join(est_phi, by = "sind")
gen_df$id <- as.factor(gen_df$id)
### model fit with B-splines basis
t1 <- Sys.time()
this_glmm <- bam(Y ~ s(sind, bs = "cr")+
                   s(id, by=phi1, bs="re")+
                   s(id, by=phi2, bs="re")+
                   s(id, by=phi3, bs="re")+
                   s(id, by=phi4, bs="re"),
                 family = binomial,
                 data=gen_df,
                 method = "fREML",
                 discrete = TRUE)
t2 <- Sys.time()
```

Estimated eigenvalues

```{r}
1/this_glmm$sp[2:5]/sqrt(n_bin)
```

Using the FPCA estimates, I was able to lower the activity but not variation. I was also not able to recreate the reversed eigenvalue problem. Also, singularity issue happened in the binnin step when using the FPCA estimates to generate data.
